# Rescue failed experiments using fractional factorial designs

Have you ever had a failed experiment? I certainly have. All the
measurements came out as 0, or the cells died, or the seeds didn't
sprout, or even the positive controls came out negative.

How can you quickly figure out what is going wrong with the minimum
amount of work? How can you quickly identify which factors are affecting
the response? The answer is fractional factorial designs.

## Example: How to bake good cookies: A fractional factorial experiment

As a familiar example, suppose we are baking cookies, but we keep
ending up with a lot of burnt cookies, a lot of soggy undercooked
cookies, or a lot of bad tasting cookies. We want to maximize the yield
of good tasting cookies. How do we quickly find out what is going wrong?

We think of 7 factors that might be affecting the yield of good tasting
cookies:

- Flour type (white or brown)
- Temperature in the oven (low or high)
- Baking soda amount (low or high)
- Time in the oven (short or long)
- Water amount (low or high)
- Sifting (sift or no)
- Butter among (low or high)

Here is a fractional factorial design to analyze cookie yield versus
these 7 factors in only 8 batches of cookies.

```{r}
cookie.data = data.frame(flour = c("brown","white","brown","white","brown","white","brown","white"),
temp = c("low","low","high","high","low","low","high","high"),
soda = c("low","low","low","low","high","high","high","high"),
time = c("long","short","short","long","long","short","short","long"),
water = c("high","low","high","low","low","high","low","high"),
sift = c("sift","sift","no","no","no","no","sift","sift"),
butter = c("low","high","high","low","high","low","low","high"),
yield = c(47, 74, 84, 62, 53, 78, 87, 60))
cookie.data[1:7] = lapply(cookie.data[1:7], factor)
cookie.data
```

Each row is a batch. The 7 columns flour, … butter, are
the factors set to their high or low values we think may affect the
yield. The last column is the yield of good cookies we get using the
specified values for the factors for each batch. For example, for the
combination of factors in row 1 (flour = brown, temp = low, etc.),
the yield is 47 good cookies.

We can quickly see which factors are important by plotting the averages
for each level of the factors with the R function `plot.design(cookie.data)`.

```{r}
plot.design(cookie.data)
```

The graph that `plot.design(cookie.data)` produces shows each factor on
the x-axis, and the average yield at each level of the factor on the y
axis. The factor "time" has the biggest difference between settings: a
yield near 80 for a short time in the oven versus a yield near 55 for a
long time in the oven. From the graph we see that time has the largest
effect on yield, followed by temperature. The other factors seem to have
little effect on yield.

Notice in the cookie data that each level of each factor gets tested in
4 batches. For example, the variable temp is low in batches 1, 2, 5,
and 6. It is high in the other 4 batches.

Batch 7, with 87 good cookies, and batch 3, with 84 good cookies, are
the best. Which factor levels are giving us the high yields? We would
like to know the average yield for each variable when it is at each
level. One way to get this is with the R function `aggregate()`.

```{r}
with(cookie.data, aggregate(yield ~ time, FUN = mean))
```

The average yield with time long is 55.5 good cookies, while the
average yield with time short is 80.75 good cookies. The difference is
80.75 - 55.5 = 25.25 good cookies. Apparently leaving them in the oven
too long may be giving us a lot of burnt cookies.

We can get the difference between the levels for each factor using the
`lm()` and `summary()` functions. (These give the results for an
analysis of variance of the cookie data.)

```{r}
lm.cookie = lm(yield ~ ., data = cookie.data)
summary(lm.cookie)
```

The output from the `lm()` and `summary()` functions confirms what we
saw in the graph, and gives us the quantitative differences: 25.25 for
time, -10.25 for temperature, and values near 1 or 2 for the other
factors.

Suppose for a moment that "sifting" was a labor-intensive or expensive
step. This experiment would be evidence that the step could be
eliminated with little effect on yield. The observed yield is actually
2.25 lower when sifting is done.

Suppose that butter was an expensive reagent. This experiment would
suggest that the butter might be used with little effect on yield.

Consider now how much we have learned with only 8 batches:

- We have identified 2 important factors.
- We have evidence that the other factors have little effect on yield.
- We have evidence that a labor-intensive step might be eliminated.
- We have evidence that the amount of an expensive reagent could be
  reduced.

How many batches would it take to get this information if we did not use
a fractional factorial design?

If we did one-variable-at-a-time (OVAT), we might do the following.

- Run 1 batch with white flour and 1 batch with brown flour = 2 batches
  for flour.
- Run 1 batch with temp low and 1 batch with temp high = 2 batches for
  temp.
- And so on for all 7 factors.

This OVAT method would require 7 factors \* 2 batches per factor = 14
batches, compared to the 8 batches for the factorial experiment design.
But OVAT has an even bigger weakness: each level of each factor is only
examined in 1 batch. In contrast, in the factorial experiment design,
each level of each factor is examined in 4 batches. These additional
replicates give much more reliable estimates of the effects, and much
greater power to detect statistical significance and get p\<0.05 for the
effect of a factor.

How do we get the p-values for the effect of the factors? For this
example, the graph indicates that temperature and time are the important
factors. We can re-run the `lm()` and `summary()` analyses using just
these two factors, which will give us the p-values.

```{r}
lm.cookie.2vars = lm(yield ~ time + temp, data = cookie.data)
summary(lm.cookie.2vars)
```

The analysis shows that both time and temperature have significant
effects (p\<0.05).

We can do a test for an interaction between time and temperature in
their effect on yield.

```{r}
lm.cookie.2vars.interaction = lm(yield ~ time * temp, data = cookie.data)
summary(lm.cookie.2vars.interaction)
```

In this case, the interaction test is not significant (p=0.73), so we
can just use the analyses of the main effects.

## Additional resources

The YouTube video "Experiments 4A -- the trade-offs when doing
half-fraction factorials" by Kevin Dunn introduces fractional factorial
designs.

A free PDF of the text *A First Course in Design and Analysis of
Experiments* by Gary Oehlert is available at
<http://users.stat.umn.edu/~gary/Book.html>.

For examples of how to do the analyses using R, see the textbook
*Design and Analysis of Experiments with R* by John Lawson (CRC Press),
and the accompanying R package `daewr`.
